---
# title: "EDA"
output: pdf_document
date: "`r Sys.Date()`"
---

```{r import, include=FALSE}
 load("projectData.RData")
```

# EDA

```{r str, echo=FALSE}
 str(data)
```

We can observe that Criminals per 100k are considered as character although they are discrete data. Thus we can convert them to integer.
Furthermore, region_name and region_code are identified as characters although they are factors. Thus we can convert them to factors.

```{r convert, include=FALSE}
data$criminals_per_100k <- as.integer(data$criminals_per_100k)
data$region_code <- as.factor(data$region_code)
data$region_name <- as.factor(data$region_name)
str(data)
attach(data)
```

## Univariate analysis

```{r summary, echo=FALSE}
# Assuming 'data' is your dataset

# Extract numerical and categorical features
numerical_features <- sapply(data, is.numeric)
categorical_features <- sapply(data, function(x) is.factor(x) | is.character(x))

# Subset data based on feature types
numerical_data <- data[, numerical_features]
categorical_data <- data[, categorical_features]

summary(numerical_data)
```
### Numerical features

In order to perform the univariate analysis for numerical predictors (here the number of ale beer houses per 100k per county, the number of public school attendants per 100k per county and the number of public whorship per 100k per county), we computed informations about the center of each distribution (the mean and the median), and about the spread (with the minimum and maximum value as well as the lower percentile 25 and the upper percentile 75).

Ajouter Standard deviation (SD), Interquartile range (IQR), Median Absolute Deviation (MAD)

While the mean is very sensitive to outliers, the median is not. Thus by comparing the two, we can have insights on the presence of outliers. We can observe that the mean and the median are quite similar relative to the spread for all three numerical features, infering against the presence of outliers.
We can assert this by computing the boxplots of the numerical features (See figure below) and observe the absence of outliers except maybe for one suspected outlier in the distribution of school attendants per county. 

In order to have a visual summary of the distributions, boxplots and histograms are displayed below.

```{r boxplots1, include=FALSE}
data$criminals_per_100k <- as.integer(data$criminals_per_100k)

# Creating a single boxplot with different colors for each feature
boxplot(ale_beer_houses_per_100k, 
        attendants_public_school_per_10k, 
        attendants_public_worship_per_2k, 
        main = "Comparison of Various Features per County",
        names = c("Beer Houses", "School Attendants", "Worship Attendants"),
        ylab = "Count per 100k")

```

```{r histograms1, include=FALSE}
# Layout to split the screen
layout(mat = matrix(c(1,2),2,1, byrow=TRUE),  height = c(3,8))
 
# Draw the boxplot and the histogram 
par(mar=c(0, 3.1, 1.1, 2.1))
boxplot(criminals_per_100k , horizontal=TRUE, xaxt="n", frame=F)
par(mar=c(4, 3.1, 1.1, 2.1))
hist(criminals_per_100k , breaks=40, border=F , main="" , xlab="value of the variable")
```
```{r boxplots2, echo=FALSE}
library(packHV)
layout(matrix(c(1, 1, 2), nrow = 1), widths = c(1.25, 1))

# Creating a single boxplot with different colors for each feature
boxplot(ale_beer_houses_per_100k, 
        attendants_public_school_per_10k, 
        attendants_public_worship_per_2k, 
        main = "Comparison of Various Features per County",
        names = c("Beer Houses", "School Attendants", "Worship Attendants"),
        ylab = "Count per 100k")
hist_boxplot(criminals_per_100k, breaks=20, ymax=6)
 
```

We can observe on both the boxplots and the histograms that the distributions are quite symmetric, while being slightly left-skewed for the Beer Houses, and the Worship attendants and slightly right-skewed for the Public School attendants. 

Since values are discrete and above 0 for all three numerical features, we can think that they belong to the Poisson distribution

#### Comparison with Poisson distribution


```{r qqplots1, include=FALSE}
par(mfrow=c(1,4)) #plt=c(0.2, 0.8, 0.2, 0.8))
qqplot(rpois(n = length(criminals_per_100k), lambda = mean(criminals_per_100k)), criminals_per_100k, xlab = "theoretical poisson")
qqplot(rpois(n = length(ale_beer_houses_per_100k), lambda = mean(ale_beer_houses_per_100k)), ale_beer_houses_per_100k, xlab = "theoretical poisson")
qqplot(rpois(n = length(attendants_public_school_per_10k), lambda = mean(attendants_public_school_per_10k)), attendants_public_school_per_10k, xlab = "theoretical poisson")
qqplot(rpois(n = length(attendants_public_worship_per_2k), lambda = mean(attendants_public_worship_per_2k)), attendants_public_worship_per_2k, xlab = "theoretical poisson")
```
Scales do not correspond exept maybe  criminals, so they are not Poisson distribued ?

```{r qqplots criminals, include=FALSE}
par(pty="s")
xlim = c(min(criminals_per_100k), max(criminals_per_100k))
qqplot(rpois(n = length(criminals_per_100k), lambda = mean(criminals_per_100k)), criminals_per_100k, xlab = "theoretical poisson",xlim = xlim, ylim = xlim)
par(pty="m")
```
Mouais

Can they be Normal ?

```{r scale, include=FALSE}
data$scaled_criminals_per_100k = (criminals_per_100k - mean(criminals_per_100k))/sd(criminals_per_100k)
data$scaled_ale_beer_houses_per_100k = (ale_beer_houses_per_100k - mean(ale_beer_houses_per_100k))/sd(ale_beer_houses_per_100k)
data$scaled_attendants_public_school_per_10k = (attendants_public_school_per_10k - mean(attendants_public_school_per_10k))/sd(attendants_public_school_per_10k)
data$scaled_attendants_public_worship_per_100k = (attendants_public_worship_per_2k - mean(attendants_public_worship_per_2k))/sd(attendants_public_worship_per_2k)

str(data)
attach(data)
```

```{r qqplots normal1, include=FALSE}
par(mfrow=c(1,2))
qqnorm(scaled_criminals_per_100k, main = "Normal QQ plot for criminals")
qqline(scaled_criminals_per_100k)
qqnorm(scaled_ale_beer_houses_per_100k, main = "Normal QQ plot for beer houses")
qqline(scaled_ale_beer_houses_per_100k)
```
```{r qqplots normal2, include=FALSE}
par(mfrow=c(1,2))
qqnorm(scaled_attendants_public_school_per_10k, main="Normal QQ plot - school attendants")
qqline(scaled_attendants_public_school_per_10k)
qqnorm(scaled_attendants_public_worship_per_100k, main = "Normal QQ plot -  worship attendants")
qqline(scaled_attendants_public_school_per_10k)
```
### Categorical features

```{r heatmap, echo=FALSE}
#library(gplots)

cont_table = table(data$region_name, data$region_code)
# Convert the table to a matrix
cont_matrix <- as.matrix(cont_table)

colors <- colorRampPalette(c("lightyellow", "blue"))(100)

# Create the heatmap
heatmap(cont_matrix, 
        col = colors, 
        scale = "none",  # To prevent scaling of values
        Rowv = NA,       # To suppress row dendrogram
        Colv = NA,       # To suppress column dendrogram
        margins = c(5, 5))  # Set margins

# Add a color key
legend("bottomleft", legend = c("0", "Above 0"), fill = c("lightyellow", "blue"), bty = "n")

```
Each region name corresponds to a region code.
We have to replace the region_code "1" of South Eastern to factor "9"

```{r summary2, echo=FALSE}
data$region_code <- factor(data$region_code, levels = c(levels(data$region_code), "9"))
data[data$region_name == "South Eastern", "region_code"] <- factor("9")

categorical_features <- sapply(data, function(x) is.factor(x) | is.character(x))
categorical_data <- data[, categorical_features]
summary(categorical_data)
```
## Bivariate analysis

```{r bivariate boxplot, echo=FALSE}
factor_colors <- c("red", "blue", "green", "orange", "purple", "yellow", "pink", "cyan") # , "magenta")

# Order region codes by median value
ordered_region <- names(sort(tapply(criminals_per_100k, region_code, median)))

# Create boxplots ordered by median
boxplot(criminals_per_100k ~ factor(region_code, levels = ordered_region), 
        main = "Criminals per 100k by Region Code",
        col = factor_colors,
        xlab = "Region Code",
        ylab = "Criminals per 100k")

```

```{r correlation, include=FALSE, warning=FALSE}
# Scatter plot of pairwise correlation between the three predictor variables:

# Selecting the predictor variables
predictor_vars <- data.frame(
  ale_beer_houses_per_100k,
  attendants_public_school_per_10k,
  attendants_public_worship_per_2k
)

# Creating the scatterplot matrix
pairs(predictor_vars, 
      main = "Pairwise Scatterplot of Predictor Variables")

```

```{r correlation2, include=FALSE, warning=FALSE}
# Customize upper panel
upper.panel<-function(x, y){
  points(x,y)
  r <- round(cor(x, y), digits=2)
  txt <- paste0("R = ", r)
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  text(0.5, 0.9, txt)
}
pairs(predictor_vars, lower.panel = NULL, 
      upper.panel = upper.panel)

```
```{r correlation3, echo=FALSE, warning=FALSE}
# Customize upper panel
upper.panel<-function(x, y){
  points(x,y, pch=19, col=factor_colors[data$region_code])
  r <- round(cor(x, y), digits=2)
  txt <- paste0("R = ", r)
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  text(0.5, 0.9, txt)
}
pairs(data[,5:7], lower.panel = NULL, 
      upper.panel = upper.panel)

```
